---
title: 'Assignment 2: Instrumental Variables'
author: "Thammarat Bunjaroenphornsuk 22037148 & Iván Felipe Moreno Guzmán 19438013"
date: "2024-05"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(haven)      # To use data sets from Stata and other programs
library(tidyverse)  # To manage data sets
library(stargazer)  # to produce well-formatted output tables
library(AER)        # to run IV regressions and obtain some data 
library(knitr)      # To knit 
options(scipen = 999)

```

# Part 1

*You want to assess the relationship between education and income (measured as consumption expenditure) in Malawi. The dataset iv dataset1 contains information on current household consumption expenditure and completed years of schooling for 70,000 individuals from 70 schools. These individuals are former students who were randomly selected from school records and contacted to receive information on their current living standards. Assume that all schools are subject to the same rules (children start school at age 6 and the highest possible number of completed school years is 12 in all schools) and that treatment effects are homogeneous. When interpreting your results, focus on causality, effect size, and statistical significance.*

```{r dataset1, results='asis'}
data <- haven::read_dta("iv_dataset1.dta")


## This first chunk of code just explore the data and the particular variables of interest
stargazer(as.data.frame(data), title="Summary",
          style = "aer", digits = 2)


p1 <- ggplot(data = data, aes(x = completed_school_years)) + 
    geom_histogram(bins = 12, fill = 'salmon', color = 'black') +
    ggtitle("")
p2 <- ggplot(data , aes(x = consexp)) + 
    geom_histogram(fill = '#2D9596', color = 'black') +
    ggtitle("")

ggpubr::ggarrange(p1,p2, ncol=2,
                  label.y = "Count" )

data_summary <- data.frame(
    Name = names(data),
    Class = sapply(data, function(x) paste(class(x),collapse = ",")),
    NAs = sapply(data, function(x) sum(is.na(x))),
    Unique_Values = sapply(data, function(x) length(unique(x))),
    stringsAsFactors = FALSE
  )

stargazer(as.data.frame(data_summary), title="Data Outlook",
          summary = F, style = "aer", digits = 2,
          column.sep.width = "2pt", 
          rownames = F)

```


## Question 1

*Run an OLS regression to investigate the effect of completed school years on consumption expenditure and interpret your results. In general, why is this approach unlikely to give you an unbiased estimate of the causal effect of educational attainment on consumption expenditure? What is an example of a likely potential source of endogeneity, and to what direction does this bias lead and why? Be as specific as possible.*

```{r firstOLS, results='asis'}
ols_reg <- lm(consexp~completed_school_years,data = data)

stargazer(ols_reg,  
          title="OLS Regression", type="latex", 
          df=FALSE,  digits = 2)

```

According to the coefficients, one extra year of education may imply an increase on average of `r ols_reg$coefficients[[2]]`. This result is significant at the 5% level. -> we can compare it with the original distribution of the variable, for example with the mean and the sd.

However, this approach can yield a biased estimate of the causal effect of educational attainment on consumption expenditure. The potential endogeneity problem arises from two sources: omitted variables and measurement errors in the independent variable. An example of the first issue is family income, which can influence both education outcomes and consumption expenditure, as higher-income families can afford better education and opportunities. For the second issue, systematic measurement errors (e.g., misreported income or consumption) can lead to biased results, as individuals might misreport to gain or avoid benefits.

*what direction does this bias lead and why? Be as specific as possible* 



## Question 2

*One of the key reasons for the low attendance and early school dropouts in Malawi are health problems. About 35% of schools were provided with a Water, Sanitation, and Hygiene (WASH) program, which has increased average student success in these schools. Based on this information, perform an instrumental variable (IV) regression to asymptotically recover the causal effect of completed school years on consumption expenditure and interpret your results.*

```{r IV, results='asis'}

W_IV <- ivreg (consexp~completed_school_years|WASHprogram ,
               data = data)  

stargazer(ols_reg, W_IV, title="Instrumental Variable- WASH",
          column.labels=c("Base OLS","IV"),
          type="latex", digits = 2)



```
In this case, when the school years are instrumented by the WASH program, the coefficient is still significant but its magnitude is lower. In this case, the effect is `r W_IV$coefficients[[2]]`

## Question 3

*Which treatment effect are you estimating in this setting and why?*

ATE because we assume treatment effect is homogeg

## Question 4

*Which assumptions are necessary for the IV approach of task 2? First state the assumptions in general terms and then apply them to the specific example of this assignment. Discuss the validity of the assumptions and, to the extent possible, perform a formal test. What bias do you expect in your IV estimates and why?*

To consider an instrumental variable as valid, it must fulfill two conditions: relevance and exclusion restriction. The first condition is that the instrument must be relevant and should explain variation on the endogenous variable. This condition is usually revised with the significance of the first stage, regressing the independent variable with the instrument and other controls, or checking the F statistic of the first stage. The second condition implies that the selected instrument has no direct relation with the dependent variable of the research, and it influence is mediated by the instrumented variable -usually this is the most difficult condition to fulfill and there are not formal methodologies to test it.

```{r IVtest}
summary(W_IV, diagnostic=T)
```
In this case, the test shows that the instrument is not weak, meaning it effectively explains the education variable. However, there may still be an endogeneity problem, as wash services are likely situated in wealthier areas. Consequently, these services might be correlated with higher income levels, which in turn could influence both education and consumption expenditure, potentially biasing the results.

## Question 5

*Now assume that the WASH program was randomly assigned. Does this change your answer with respect to the validity of the IV assumptions? If yes, how?*

destroys relevance, correrlation z and x dissapears 

## Question 6

*You know that sex and age of former students are important predictors of income, so you decide to control for them in your regressions. Repeat the regressions from task 1 and task 2 with these additional controls and interpret your results. Do your results change compared to before? Why?Why not?*

```{r,  results='asis'}

ols_reg_control <- lm(consexp~completed_school_years +age+sex,
                      data = data)

W_IV_control <- ivreg (consexp~completed_school_years+age+sex|WASHprogram+age+sex ,
               data = data)

stargazer(ols_reg, W_IV,  
          ols_reg_control,W_IV_control,
          title="With Age and Sex controls",
          type="latex", digits = 2)


```


# Part 2

*You are evaluating a randomized controlled trial (RCT) in 250 primary schools in India that aims to tackle micronutrient deficiencies. As intervention, an NGO provided schools with a “Lucky Iron Leaf”, a small iron cooking tool that infuses your meals with iron to help prevent and treat iron deficiency. Your data is provided in the dataset iv dataset2. Your research goal is to investigate the effect of the iron ingot provision on school performance. For this purpose, of the 250 schools, 125 were randomly selected to receive “Lucky Iron Leaf”. At the end of the year, student performance was assessed using a standardized test. When looking at the data, you find out that not all schools in the treatment group actually used the iron ingot, and you suspect that treatment effects may be heterogeneous.*\

we cannot assumme homo - ITT

## Question 1

*Calculate the Intention-to-Treat effect (ITT) of iron ingot provision on student test scores and interpret your result. How does this effect differ from the conventional ATE? Why may the ITT still be of interest in the described example?*

lecture 5:
"One option is measure the ITT, i.e. the causal effect of treatment
assignment regardless of actual participation (does not require
further assumptions due to random assignment)"

evaluate program - not just giving b

```{r}
data2 <- haven::read_dta("iv_dataset2.dta")
itt_iron <- lm(testscores~ironleaf_provided,
                      data = data2)

stargazer(itt_iron, title="ITT of iron ingot provision",
          type="latex", digits = 2)
```


Effect is on the ones who receive the treatment 

## Question 2

*Based on the estimated ITT, derive a consistent estimate for the causal effect of the treatment on compliers and interpret your result. For this, please do not use any IV regression function. Instead, calculate the effect using your ITT estimate with the Indirect Least Squares approach. Standard errors can be ignored for this task. Briefly comment on the results.*

```{r}

# reduced form and first stage

FS_A <- lm(ironleaf_used~ironleaf_provided,data = data2) # Compliance rate
RF_A <- lm(testscores~ironleaf_provided,data = data2) # ITT

coef(RF_A)[2] / coef(FS_A)[2] # reduced form / first stage

```


## Question 3

*Replicate your results from task 2 using a “manual” Two-Stage-Least-Squares approach (2SLS), i.e. do not use any IV regression functions. Standard errors can be ignored for this task.*

```{r}
first_stage <- lm(ironleaf_used~ironleaf_provided, data = data2)
second_stage <- lm(data2$testscores~first_stage$fitted.values)

stargazer(second_stage, title="Two-Stage-Least-Squares approach (2SLS)", 
          digits = 2,type="latex")

```


lecture 5 - slide 7 

## Question 4

*Which assumptions must be fulfilled for your estimate from task 2 and task 3 to be a local average treatment effect? Under which assumption is LATE identical to ATE? Please be specific to the analysis at hand!*


